% Blueprint content for the infinitary logic formalization

% Load generated node definitions from LeanArchitect
\input{../../.lake/build/blueprint/library/InfinitaryLogic}
\input{../../.lake/build/blueprint/library/InfinitaryLogicConditional}

\chapter{Infinitary Logic}

This project formalizes results in infinitary logic in Lean~4 using Mathlib.
The primary focus is the logic $\Lomegaone$, which extends first-order logic
with countable conjunctions and disjunctions, though the formalization
treats the more general $\Linf$ (allowing arbitrary conjunctions and
disjunctions) where possible.

\section{Main Results}

The formalization establishes ten headline theorems:
\begin{itemize}
  \item \textbf{Scott characterization.}
    Every countable structure is characterized up to isomorphism by
    a sentence of $\Lomegaone$ (Theorem~\ref{thm:scott-characterizes}).
  \item \textbf{Scott rank below $\omegaone$.}
    The Scott rank of any countable structure is a countable ordinal
    (Theorem~\ref{thm:scottRank-lt-omega1}).
  \item \textbf{Karp's theorem.}
    Two structures are $\Linf$-equivalent if and only if there exists
    a back-and-forth system between them (Theorem~\ref{thm:karp-theorem}).
  \item \textbf{Model existence.}
    Every countable consistent set of sentences of $\Lomegaone$ in a countable
    language has a countable model (Theorem~\ref{thm:model-existence}).
  \item \textbf{Countable refinement hypothesis.}
    The back-and-forth hierarchy on any countable structure stabilizes
    in countably many steps (Theorem~\ref{thm:CRH}).
  \item \textbf{Silver's theorem and the Silver--Burgess dichotomy.}
    A Borel equivalence relation on a Polish space has countably many classes
    or a perfect set of pairwise-inequivalent points; on a standard Borel space
    the quotient has size $\leq\aleph_0$ or exactly $2^{\aleph_0}$
    (Theorem~\ref{thm:silver-burgess-proved}).
  \item \textbf{Morley counting.}
    An $\Lomegaone$ sentence has $\leq\aleph_1$ or exactly $2^{\aleph_0}$
    isomorphism classes of countable models (Theorem~\ref{thm:morley-counting}).
  \item \textbf{The Morley--Hanf theorem.}
    $\beth_{\omegaone}$ is a Hanf bound for every $\Lomegaone$ sentence,
    over an arbitrary language with no side hypotheses
    (Theorem~\ref{thm:morley-hanf}).
  \item \textbf{The exact Hanf number.}
    $\mathrm{Hanf}(\Lomegaone) = \beth_{\omegaone}$: the Morley--Hanf bound is
    sharp, by Marker's Exercise~5.3 beth ladder over the von Neumann hierarchy
    (Theorem~\ref{thm:hanf-exact}).
  \item \textbf{Small models of every infinite size.}
    A sentence with arbitrarily large models has, at every infinite $\kappa$, a
    model of size exactly $\kappa$ realizing only countably many complete
    $\Lomegaone$-types (Theorem~\ref{thm:small-models}).
\end{itemize}

\section{Back-and-Forth Equivalence}

Back-and-forth equivalence at ordinal levels, following Karp~\cite{karp1965}.

\inputleannode{def:BFEquiv}

\section{Scott Formulas and Sentences}

Scott formulas and their characteristic property, following
Marker~\cite[Ch.~2]{marker2016}.

\inputleannode{def:scottFormula}

\inputleannode{thm:scottFormula-iff}

\section{Self-Stabilization}

\inputleannode{thm:self-stabilization}

\section{Scott Sentence}

\inputleannode{def:stabilization-ordinal}

\inputleannode{def:scott-sentence}

\inputleannode{thm:scott-characterizes-of}

\chapter{Countable Refinement}

The countable refinement hypothesis and its consequences for Scott rank.

\section{Countable Refinement Hypothesis}

\inputleannode{thm:CRH}

\emph{Note:} Despite its name, the ``Countable Refinement Hypothesis'' is fully
proved in this formalization.  The name is historical: it was originally an
open conjecture about the countability of refinement ordinals.  The proof here
proceeds via self-stabilization and transfinite induction.

This countability result drives the downstream stabilization argument: since
only countably many ordinals can witness a strict refinement, the hierarchy
must stabilize below~$\omegaone$.

\inputleannode{thm:scott-characterizes}

\section{Scott Rank and Height}

\inputleannode{def:scottRank}

\inputleannode{thm:scottRank-lt-omega1}

\inputleannode{def:scottHeight}

\inputleannode{thm:scottHeight-lt-omega1}

\chapter{Karp's Theorem}

Karp's theorem characterizes $\Linf$-equivalence via back-and-forth
systems, following Karp~\cite{karp1965}.

\inputleannode{def:potential-iso}

\inputleannode{def:linf-equiv}

\inputleannode{thm:karp-theorem}

\chapter{Model Existence}

Model existence and completeness for $\Lomegaone$, following
Keisler~\cite{keisler1971}.

\inputleannode{thm:model-existence}

\inputleannode{thm:karp-completeness}

\section{Omitting Types}

\inputleannode{def:omits-type}

\inputleannode{thm:omitting-types}

\chapter{L\"owenheim--Skolem}

The downward L\"owenheim--Skolem theorem for $\Lomegaone$.

\inputleannode{def:naming-function-with-constants}

\inputleannode{thm:downward-ls}

\inputleannode{thm:downward-ls-theory}

\chapter{Admissible Fragments}

Compactness and completeness for countable admissible fragments of
$\Lomegaone$, following Barwise~\cite{barwise1975}.  Background on the Nadel bound
for admissible fragments can be found in Nadel~\cite{nadel1974}.

\inputleannode{def:admissible-fragment}

\inputleannode{thm:barwise-compactness}

\inputleannode{thm:barwise-completeness-ii}

\chapter{Hanf Numbers}

Hanf numbers measure how large a model a sentence must have before
it has arbitrarily large models.

\inputleannode{def:arb-large-models}

\inputleannode{def:hanf-bound}

\inputleannode{def:hanf-number}

\inputleannode{thm:hanf-existence}

\section{Morley--Hanf Bound}

\emph{Status: proved, unconditionally.}  The theorem \texttt{morley\_hanf}
(\ref{thm:morley-hanf}) states that $\beth_{\omegaone}$ is a Hanf bound for
every $\Lomegaone$ sentence, over an arbitrary language with no side
hypotheses.  No extraction is consumed at all: any injective
$\mathbb{N}$-sequence of the source is already fully indiscernible on the
Morley seed (\texttt{morleySeed\_indiscernibleOn}).  The single substantive
input is realizability of the seed's Ehrenfeucht--Mostowski tail-template
theory over every target order
(\texttt{morleySeedTailTemplateRealizable\_holds}).  The latter is a
Henkin-style construction in the spirit of Marker's presentation of Morley's
omitting-types argument: an $\omega$-stage completion of the countable
\emph{schema} sentence universe --- deciding each lifted template sentence and
pinning a disjunct for every positive countable disjunction and a refuted
conjunct for every negative countable conjunction --- whose quotient term model
carries a fully indiscernible sequence of Henkin constants satisfying a
restricted truth lemma.  Symbol countability is removed by running the
construction in the sublanguage generated by the sentence's own countably many
symbols and expanding the resulting model back to the full language.

The historical conditional forms are retained: the original
\texttt{morley\_hanf\_of\_transfer} (\ref{thm:morley-hanf-transfer}) with the
bundled transfer hypothesis, and the split bridge
\texttt{hasArbLargeModels\_of\_restricted\_extraction}
(\ref{thm:morley-hanf-restricted}) consuming the (since refuted-in-ZFC)
full-indiscernibility extraction residual plus a compactness oracle.

\inputleannode{thm:morley-hanf}

The milestones of the schema construction behind the seed-realizability input:

\inputleannode{thm:schema-truth-lemma}

\inputleannode{thm:schema-seq-indiscernible}

\inputleannode{thm:morley-seed-realizable}

\inputleannode{thm:morley-hanf-transfer}

\inputleannode{thm:morley-hanf-restricted}

\section{Sharpness: the Exact Hanf Number}

\emph{Status: proved.}  The Morley--Hanf bound is exact:
$\mathrm{Hanf}(\Lomegaone) = \beth_{\omegaone}$
(\texttt{Lomega1omegaHanfNumber\_eq\_beth\_omega1}, \ref{thm:hanf-exact}).
The upper half (\ref{thm:hanf-upper-global}) is immediate from the Morley--Hanf
theorem.  The lower half (\ref{thm:hanf-lower-global}) is Marker's
Exercise~5.3: for each $\alpha < \omegaone$ a single $\Lomegaone$ sentence
over countably many constants, unary levels indexed by the well order
$\alpha+2$, and one binary extensional edge relation has models of size
exactly $\beth_{\alpha+1}$ at the maximum (\ref{thm:hanf-ladder-stage}).
The maximal model is the von Neumann hierarchy: the levels are interpreted as
the stages $V_{\omega+\beta}$, whose cardinalities Mathlib already computes,
shrunk to a $\mathrm{Type}$-$0$ carrier; the upper bound is a well-founded
induction on the level order with powerset steps at successors and countable
unions at limits.  The successor-cofinal supremum
$\sup_{\alpha<\omegaone}\beth_{\alpha+1} = \beth_{\omegaone}$ closes the
argument.

\inputleannode{thm:hanf-upper-global}

\inputleannode{thm:hanf-ladder-stage}

\inputleannode{thm:hanf-lower-global}

\inputleannode{thm:hanf-exact}

\section{Small Models}

\emph{Status: proved.}  Marker's Theorem~11.2
(\texttt{exists\_small\_model\_of\_hasArbLargeModels}, \ref{thm:small-models}): a sentence
with arbitrarily large models has models of every infinite size realizing only countably many
complete $\Lomegaone$-types.  The witnesses are the local Ehrenfeucht--Mostowski quotients of
the Morley-seed schema source over highly order-transitive skeletons (linear ordered fields,
supplied at every infinite cardinality by lexicographic Hahn-series subfields); the quotient is
equivariant under order automorphisms of the skeleton, tuples are classified by countably many
compressed term codes, and located term codes compute the carrier cardinality exactly.
Arbitrary languages reduce along a uniform collapsing hom.

\inputleannode{thm:small-models}

\chapter{Counting Models}

Counting the number of models of a Scott sentence up to isomorphism.
The theorem \texttt{morley\_counting\_dichotomy} (a counting-models
dichotomy in the style of Morley) is formalized in Lean but is not
included as a blueprint node; it serves as a schematic boundary result.

\emph{Status: schematic.}  The dichotomy statement quantifies over
cardinals and does not correspond to a single blueprint-trackable proof
obligation.

\inputleannode{thm:stabilization-bound-iso}

\inputleannode{thm:bounded-scott-height-iso}

\chapter{Proof System for Admissible Fragments}

A proof system for admissible fragments of $\Lomegaone$, following Barwise~\cite{barwise1975}.

\inputleannode{def:derivable}

\inputleannode{def:a-consistent}

\inputleannode{thm:proof-system-soundness}

\inputleannode{def:full-barwise-fragment}

\inputleannode{thm:consistency-property-full-fragment}

\inputleannode{thm:barwise-completeness-ii-syntactic}

\chapter{Descriptive Set Theory}

Borel complexity of satisfaction, back-and-forth equivalence, and isomorphism
on the coding space for countable structures.

\inputleannode{def:structure-space}

\inputleannode{thm:satisfaction-borel}

\inputleannode{thm:bfequiv-borel}

\inputleannode{thm:iso-borel}

\section{Standard Borel Structure}

The structure space carries the product topology, which makes it a Polish space
when the language has countably many relation symbols.

\inputleannode{thm:structure-space-polish}

The model class of any $\Lomegaone$ sentence inherits a standard Borel structure.

\inputleannode{thm:models-clopenable}

\inputleannode{thm:models-standard-borel}

\section{Counting Models}

The Silver--Burgess dichotomy, taken as a hypothesis, yields a counting theorem
for coded $\mathbb{N}$-models with bounded Scott height.

\inputleannode{def:silver-burgess-dichotomy}

\inputleannode{thm:silver-burgess-proved}

\inputleannode{def:iso-setoid}

\inputleannode{thm:counting-dichotomy}

\inputleannode{thm:finite-carrier-iso-borel}

\inputleannode{thm:counting-all-countable}

\inputleannode{thm:morley-counting}

The combined theorem extends the coded $\mathbb{N}$-model result to all countable
models by adding finite-carrier tiers (isomorphism on $\operatorname{Fin} n$
structures is Borel via the orbit of $\operatorname{Sym}(n)$).  Bridge theorems
(\texttt{codeModel}, \texttt{iso\_of\_codeModel\_eq}, \texttt{codeModel\_surjective})
establish that \texttt{AllCodedIsoClasses} faithfully represents all isomorphism
classes of countable models.  The Silver--Burgess dichotomy
(\texttt{silverBurgessDichotomy}) and Silver's theorem
(\texttt{gandy\_harrington\_for\_relation}) are \emph{fully proved}, via
Miller's classical category route (Kuratowski--Ulam, Mycielski, the
Kechris--Solecki--Todorcevic $G_0$-dichotomy); \texttt{morley\_counting} is
therefore unconditional, with axioms exactly
\texttt{[propext, Classical.choice, Quot.sound]}.

\bibliographystyle{amsalpha}
\bibliography{refs}
